---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
 This is Peer Assessment -1 Work for 
 Reproduciable Research course on Coursera
 
 Created by: Bhaksar Bandaru

 Date: 17th May 2015


## Loading and preprocessing the data

Now we load the data 

```{r loaddata, echo=TRUE,results='hide'}
require (reshape2)
require (ggplot2)
require (dplyr)

unzip("activity.zip", exdir = ".",  overwrite=TRUE)
activitydf <- read.csv("activity.csv")

```
 

Now we do initial explrotation of the data

```{r datacheck, echo=TRUE,results='markup'}

names(activitydf) 

dim (activitydf)

head(activitydf,3)

tail (activitydf,3)

## now check the levels of the date

unique(activitydf$date)


##check for any missing values
summary(activitydf$steps)

## now the percentage of the missing values are 
mean(is.na(activitydf$steps))

## The percentage of misssing values is statistically small 

```
##### The percentage of misssing values is statistically small 13%

## What is mean total number of steps taken per day?

Now calculate the Daily activity steps 

```{r dailyactivity, echo=TRUE,results='markup'}

## Now we would calculate the daily activity totals with out missing value 

sumbydate <- aggregate(steps ~ date, data= activitydf, sum)

dim (sumbydate)

head(sumbydate,3)

tail (sumbydate,3)

##now the Mean and Median values
## mean value is 
mn <- mean (sumbydate$steps)
mn 

##median value is 
md <- median (sumbydate$steps)
md


```
The mean value is `r mn` and the median value is `r md` 

now plot the histogram with the mean value

```{r firsthistogram, echo=FALSE,fig.height=4,fig.width=5}


hist (sumbydate$steps, col = "lightblue", xlab = "Total Activity steps per day", main ="Histogram of Total Steps per Day" )

abline (v= mn, col = "red", lwd ="2")


```

## What is the average daily activity pattern?

Now we want to calculate the average dauly activity in the time series of the interval

```{r dailyaverage, echo=TRUE,results='markup'}

## Now we would calculate the mean daily activity with the time interval 

meandaily <- aggregate(steps ~ interval, data= activitydf, mean)

dim (meandaily)

head(meandaily,3)

tail (meandaily,3)

## Now get the max value of the average and the time period

maxvalue <- filter (meandaily, steps == max(steps))
maxsteps <- maxvalue$steps
maxint <- maxvalue$interval
## Maximum average steps
maxsteps
## Interval at which the max average steps daily
maxint

## this provides the real conversion to Hours and Minutes
maxhr <- round (maxint/100)
maxhr
maxmin <- (maxint-maxhr*100)
maxmin


```
The maximum average value is `r maxsteps` and it happens at `r maxhr` Hours and `r maxmin` Minutes

Now create the time series plot
 
```{r firstplot, echo=FALSE,fig.height=4,fig.width=5}


plot (meandaily, type ="l", col = "blue", xlab = "Time interval (in 5 Min) ", ylab = "Average steps per Day", main ="Time series plot of Average Steps per Day" )



```


## Imputing missing values

##### Now here we do the Imputing the Missing values in the dataset

```{r imputingdata, echo=TRUE,results='markup'}

## First we will Split the data frame by the dates
bydate <- activitydf$date

listactivity <- split (activitydf , bydate)

## Now add the mean value per each interval daily into this temporary data set
## As we would use the average interval value to impute the NA values 
listactivity <- lapply( listactivity, transform, intmean = meandaily$steps)

##now unsplit the data set 
activitydf1 <- unsplit (listactivity, bydate)

## Replace the NA values with the average value per interval
activitydf1 <- mutate(activitydf1, steps = ifelse(is.na(steps), intmean,steps))

## now we get the clean data with all the NA values filled 
## and cleaned with removing the temp  variables like 'intmean'
activitydf1 <- select (activitydf1 , -intmean)

##let us check the values of modified data frame
names (activitydf1)

dim(activitydf1)

head (activitydf1,3)

tail (activitydf1,3)

summary (activitydf1$steps)

## No NA values in the steps

## Now calculate the total steps per day 
totalperday <- aggregate (steps ~ date, data = activitydf1, sum)
dim (totalperday)

head(totalperday,3)

tail (totalperday,3)

##now the Mean and Median values
## mean value is 
newmean <- mean (totalperday$steps)
newmean 

##median value is 
newmed <- median (totalperday$steps)
newmed
## Note that the mean and median are same as there are no missing values
## Also note that the mean is not that much different from the old value

```

###### Note: 

The old mean value is `r mn` and the new value is `r newmean`. The values are same

The old Median value is `r md` and the new value is `r newmed`.

This indicates not much difference due to missing values as the percentage is small.

Now we would show the histogram with new values and the mean value line.

```{r secondhistogram, echo=FALSE,fig.height=4,fig.width=5}


hist (totalperday$steps, col = "lightgreen", xlab = "Total Activity steps per day", main ="New Histogram of Total Steps per Day" )

abline (v= newmean, col = "red", lty ="dotted")


```


## Are there differences in activity patterns between weekdays and weekends?
