---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
 This is Peer Assessment -1 Work for 
 Reproduciable Research course on Coursera
 
 Created by: Bhaksar Bandaru

 Date: 17th May 2015


## Loading and preprocessing the data

Now we load the data 

```{r loaddata, echo=TRUE,results='hide'}
require (reshape2)
require (ggplot2)
require (dplyr)

unzip("activity.zip", exdir = ".",  overwrite=TRUE)
activitydf <- read.csv("activity.csv")

```
 

Now we do initial explrotation of the data

```{r datacheck, echo=TRUE,results='markup'}

names(activitydf) 

dim (activitydf)

head(activitydf,3)

tail (activitydf,3)

## now check the levels of the date

unique(activitydf$date)


##check for any missing values
summary(activitydf$steps)

## now the percentage of the missing values are 
mean(is.na(activitydf$steps))

## The percentage of misssing values is statistically small 

```
##### The percentage of misssing values is statistically small 13%

## What is mean total number of steps taken per day?

Now calculate the Daily activity steps 

```{r dailyactivity, echo=TRUE,results='markup'}

## Now we would calculate the daily activity totals with out missing value 

sumbydate <- aggregate(steps ~ date, data= activitydf, sum)

dim (sumbydate)

head(sumbydate,3)

tail (sumbydate,3)

##now the Mean and Median values
## mean value is 
mn <- mean (sumbydate$steps)
mn 

##median value is 
md <- median (sumbydate$steps)
md


```
The mean value is `r mn` and the median value is `r md` 

now plot the histogram with the mean value

```{r firsthistogram, echo=FALSE,fig.height=4,fig.width=5}


hist (sumbydate$steps, col = "lightblue", xlab = "Total Activity steps per day", main ="Histogram of Total Steps per Day" )

abline (v= mn, col = "red", lwd ="2")


```

## What is the average daily activity pattern?

Now we want to calculate the average dauly activity in the time series of the interval

```{r dailyaverage, echo=TRUE,results='markup'}

## Now we would calculate the mean daily activity with the time interval 

meandaily <- aggregate(steps ~ interval, data= activitydf, mean)

dim (meandaily)

head(meandaily,3)

tail (meandaily,3)

## Now get the max value of the average and the time period

maxvalue <- filter (meandaily, steps == max(steps))
maxsteps <- maxvalue$steps
maxint <- maxvalue$interval
## Maximum average steps
maxsteps
## Interval at which the max average steps daily
maxint

## this provides the real conversion to Hours and Minutes
maxhr <- round (maxint/100)
maxhr
maxmin <- (maxint-maxhr*100)
maxmin


```
The maximum average value is `r maxsteps` and it happens at `r maxhr` Hours and `r maxmin` Minutes

Now create the time series plot
 
```{r firstplot, echo=FALSE,fig.height=4,fig.width=8}


p1 <- ggplot(meandaily, aes(interval, steps))

g1 <- (p1 + geom_line()  + labs(title = "Average Steps per Day") + labs (x = "Time Interval") + labs(y = "Average steps per Day"))
print (g1)


```


## Imputing missing values

##### Now here we do the Imputing the Missing values in the dataset

```{r imputingdata, echo=TRUE,results='markup'}

## First we will Split the data frame by the dates
bydate <- activitydf$date

listactivity <- split (activitydf , bydate)

## Now add the mean value per each interval daily into this temporary data set
## As we would use the average interval value to impute the NA values 
listactivity <- lapply( listactivity, transform, intmean = meandaily$steps)

##now unsplit the data set 
activitydf1 <- unsplit (listactivity, bydate)

## Replace the NA values with the average value per interval
activitydf1 <- mutate(activitydf1, steps = ifelse(is.na(steps), intmean,steps))

## now we get the clean data with all the NA values filled 
## and cleaned with removing the temp  variables like 'intmean'
activitydf1 <- select (activitydf1 , -intmean)

##let us check the values of modified data frame
names (activitydf1)

dim(activitydf1)

head (activitydf1,3)

tail (activitydf1,3)

summary (activitydf1$steps)

## No NA values in the steps

## Now calculate the total steps per day 
totalperday <- aggregate (steps ~ date, data = activitydf1, sum)
dim (totalperday)

head(totalperday,3)

tail (totalperday,3)

##now the Mean and Median values
## mean value is 
newmean <- mean (totalperday$steps)
newmean 

##median value is 
newmed <- median (totalperday$steps)
newmed
## Note that the mean and median are same as there are no missing values
## Also note that the mean is not that much different from the old value

```

###### Note: 

The old mean value is `r mn` and the new value is `r newmean`. The values are same

The old Median value is `r md` and the new value is `r newmed`.

This indicates not much difference due to missing values as the percentage is small.

Now we would show the histogram with new values and the mean value line.

```{r secondhistogram, echo=FALSE,fig.height=4,fig.width=5}


hist (totalperday$steps, col = "lightgreen", xlab = "Total Activity steps per day", main ="New Histogram of Total Steps per Day" )

abline (v= newmean, col = "red", lty ="dotted")


```


## Are there differences in activity patterns between weekdays and weekends?

###### Now we calculate the weekdays and weekends average actvity trends


```{r weekdayandweekenddata, echo=TRUE,results='markup'}

## First we willis determine the day is weekday or weekend based on data variable

## create new revised data frame with the additional variable to set the day type

activitydf2 <- mutate ( activitydf1 , day = ifelse( ((weekdays(as.Date(date)) == "Saturday")|(weekdays(as.Date(date)) ==  "Sunday")), "weekend", "weekday"))

## Now segregate the data for the week days
activitywkday <- filter(activitydf2, day == "weekday")

## and for the weekend days
activitywkend <- filter (activitydf2,day == "weekend")

## now we do average the activity for the weekday and weekend 
meanwkday <- aggregate (steps ~ interval, data = activitywkday, mean)
meanwkend <- aggregate (steps ~ interval, data = activitywkend, mean)

## Data sanity check for the weekday averages
names (meanwkday)
dim (meanwkday)

head (meanwkday,3)
tail (meanwkday,3)

##Data sanity check for the weekend averages
names (meanwkend)
dim (meanwkend)

head (meanwkend,3)
tail (meanwkend,3)


## Now get the max value of the average and the time period for weekday activity

maxwkday <- filter (meanwkday, steps == max(steps))
maxstepswkd <- maxwkday$steps
maxintwkd <- maxwkday$interval
## Maximum average steps on weekday
maxstepswkd
## Interval at which the max average steps  on weekday
maxintwkd

## this provides the real conversion to Hours and Minutes
maxhrwkd <- round (maxintwkd/100)
maxhrwkd
maxminwkd <- (maxintwkd-maxhrwkd*100)
maxminwkd

## Now get the max value of the average and the time period for weekend activity

maxwkend <- filter (meanwkend, steps == max(steps))
maxstepswke <- maxwkend$steps
maxintwke <- maxwkend$interval
## Maximum average steps on weekday
maxstepswke
## Interval at which the max average steps  on weekday
maxintwke

## this provides the real conversion to Hours and Minutes
maxhrwke <- round (maxintwke/100)
maxhrwke
maxminwke <- (maxintwke-maxhrwke*100)
maxminwke


##Now mmerge the two mean data variables into one data frame
 
meanwkday <- mutate(meanwkday, day = "weekday")
meanwkend <- mutate(meanwkend, day = "weekend")
activitydf3 <- rbind(meanwkday, meanwkend)
## Now make the factor for the day for weekday and weekend
activitydf3 <- mutate (activitydf3, day = factor(day))


##let us check the data frame
names (activitydf3)

dim(activitydf3)

head (activitydf3,3)

tail (activitydf3,3)


```

##### Note: 

We could see that the weekday on the average the max steps activity happens 
at `r maxhrwkd` Hr and `r maxminwkd`.

Where as on the weekend on the average the max steps activity happens
at `r maxhrwke` Hr and  `r maxminwke`. Which indicates 1 Hr late.


##### Now we would show the plot of the average values per weekdays and weekends.

```{r secondplot, echo=FALSE,fig.height=4,fig.width=8}

p2 <- ggplot(activitydf3, aes(interval, steps))

g2 <- (p2 + geom_line(aes(color = day)) + facet_wrap (~ day, nrow = 2, ncol =1) + labs(title = "Average Steps/Day comparison for Weekday & Weekend") + labs (x = "Time Interval") + labs(y = "Average steps perday"))
print (g2)


```


##### Note: 


##### The time series plot indicates that on average steps per week day is  relatively higher activity compared at the start of the day may be office hours and where as the average trend on the weekend is much smoother.